---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(MF)
library(tidyverse)
data(output1)
data(output2)
```


## SOFTWARE NEEDED TO RUN MF IN R
- Required: [R](https://cran.r-project.org/)
- Suggested: [RStudio IDE](https://www.rstudio.com/products/RStudio/#Desktop)

## HOW TO RUN MF:

```{r eval=FALSE, echo=FALSE}
## install the latest version from github
install.packages('devtools')
library(devtools)
install_github('yayunteng/MF')
## import packages
library(MF)
```


Here are four main functions we provide in this package : 

- **MF_measures** : Computing quantification or decomposition of multi-fuctionality measures when all functions are assumed to be independent, or when the correlation between any two functions is adjusted for. In the latter case, our MF value is obtained based on an integrated measure by considering all possible threshold levels.
- **ggmf** : Visualizing the output from the function `MF_measures`


## Multi-functionality: MF_measures()

We first describe the main function `MF_measures()` with default arguments: 

```{r eval=FALSE}
MF_measures <- function(data, q = c(0,1,2), fun_cols = 1:ncol(data),
                        ecosystem = "single", species_col = NULL,
                        is_normalized = TRUE, negative = NULL, is_zero_miss = TRUE)
```

The arguments of this function are briefly described below, and will be explained in more details by illustrative examples in later text. This main function computes quantification or decomposition multi-fuctionality of order q  when all functions are assumed to be independent or when the correlation is considered.

<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr class="header">
<th align="center">Argument</th>
<th align="left">Description</th>

</tr>
</thead>
<tbody>
<tr class="odd">
<td align="center"><code>data</code></td>
<td align="left">
(a) For <code>is_normalized = TRUE</code>, data can be input as a vector of functions (for a single assemblage), <code>matrix/data.frame</code> (assemblages by functions). In this case, the data input must be normalized between 0 and 1 already. \cr
(b) For <code>is_normalized = FALSE</code>, data can only be input as a matrix/data.frame (multiple assemblages by functions).</td>

</tr>
<tr class="even">
<td align="center"><code>q</code></td>
<td align="left">a numerical vector specifying the diversity orders. Default is 0, 1 and 2..</td>

</tr>
<tr class="odd">
<td align="center"><code>fun_cols</code></td>
<td align="left">the order number of the columns which be used as the function variable..</td>

</tr>
<tr class="even">
<td align="center"><code>ecosystem</code></td>
<td align="left"><code>'single'</code> or <code>'multiple'</code>. Determine MF measures in a single ecosystem or multiple ecosystems.</td>

</tr>
<tr class="odd">
<td align="center"><code>species_col</code></td>
<td align="left">(required only when <code>ecosystem = 'multiple'</code>), name of the column to be used to calculate the species diversity.</td>

</tr>
<tr class="even">
<td align="center"><code>is_normalized</code></td>
<td align="left">a logical variable to represent whether the data input is normalized between 0 and 1, default is <code>TRUE</code>.</td>

</tr>
<tr class="odd">
<td align="center"><code>negative</code></td>
<td align="left">(required only when <code>is_normalized = FALSE</code>), name of columns to be normalized negatively.</td>

</tr>
<tr class="even">
<td align="center"><code>is_zero_miss</code></td>
<td align="left">(required only when <code>is_normalized = FALSE</code>), a logical variable to determine whether the zero in data represents no value or not, default is <code>TRUE</code>.</td>



</tbody>
</table>

This function returns an `"mf"` object which can be further used to make plots using the function `ggmf()` to be described below. 



### Example Data
A forest tree species and ecosystem function data is included in `MF` package for illustration. In the data, a total of 209 plots were established in mature forests in six countries, representing six major European forest types: boreal forest (Finland); hemi-boreal (Poland); temperate deciduous (Germany); mountainous deciduous (Romania); thermophilous deciduous (Italy); and Mediterranean mixed (Spain). In each country, three to five common tree species were selected for the species pool; species richness in 209 plots consisted of one to five species sampled from the species pool. A total of 26 ecosystem functions or properties were measured in each plot.

For this data, the following commands display how to compute multi-functionality and corresponding decomposition in multiple ecosystems .

Run the following code to view normalized forest data: (Here we only show the first ten rows of the first two ecosystem functions)

```{r eval=FALSE}
data("Europe_Forest")
Europe_Forest
```

```{r echo=FALSE}
data("Europe_Forest")
head(cbind(Europe_Forest[1:3], round(Europe_Forest[4:5], 3)), 10)
```


### Computing Multi-functionality Via Example

#### When data is already normalized.
For example, given a normalized data, we run the `mf()` function without setting `negative` and `is_zero_miss` to compute uncorrelated and correlated multi-functionality. (Here we only show the first six outputs)

```{r eval=FALSE}
data("Europe_Forest")
output1 <- MF_measures(data = Europe_Forest, fun_cols = 4:29, ecosystem = 'single')
output1
```

```{r echo=FALSE}
# output1 <- MF_measures(data = Europe_Forest, fun_cols = 4:29, ecosystem = 'single')
tmp <- output1
tmp$qMF <- round(tmp$qMF, 3)
head(tmp, 6)
```

The output contains ID of plot (`plotid`), species richness (`target_species_richness`), composition of species (`composition`), country which the plot belong to (`Country`), `Type` (uncorrelated or correlated), the diversity order (`Order.q`) and multi-functionality (`qMF`). 

#### When data is not normalized.
Given a raw data, we run the `mf()` function with setting `negative`, names of functions with smaller value as better, and `is_zero_miss` to compute uncorrelated and correlated multi-functionality. 

```{r eval=FALSE}
data("Europe_Forest_raw")
output2 = MF_measures(data = Europe_Forest_raw, fun_cols = 4:29, ecosystem = 'single',
                      is_normalized = FALSE, negative = c("soil_cn_ff_10","wue"), is_zero_miss = TRUE)
output2
```

### Computing Decomposition of Multi-functionality In Multiple Ecosystems Via Example
For example, we run the `MF_measures()` function with forest data to compute uncorrelated and correlated multi-functionality relationships across spatial scales. (Here we only show the first six outputs)

```{r eval=FALSE}
data("Europe_Forest")
output3 = MF_measures(data = Europe_Forest, fun_cols = 4:29,
                      ecosystem = 'multiple', species_col = "composition")
output3
```


```{r echo=FALSE}
# output2 <- MF_measures(data = Europe_Forest, fun_cols = 4:29,
#                       ecosystem = 'multiple', species_col = "composition")
tmp <- output2
tmp[,6:14] <- round(tmp[,6:14], 3)
head(tmp[1:7], 6)
```
```{r echo=FALSE}
head(tmp[8:14], 6)
```

Each row of the output represents the decomposition of species richness and multi-functionality across two plots. The output contains ID (`plotid`), species richness (`target_species_richness`), composition of species (`composition`), `Country` of two plots, diversity order (`Order.q`), decomposition of species richness(`Species_gamma`, `Species_alpha`, `Species_beta`), uncorrelated multi-functionality(`MF_gamma`, `MF_alpha`, `MF_beta`) and correlated multi-functionality(`MF_gamma_cor`, `MF_alpha_cor`, `MF_beta_cor`).


## Graphic Display: ggmf()

The function `ggmf()`, which extends `ggplot2` to the `"mf"` object with default arguments, is described as follows: 

```{r eval=FALSE}
ggmf(output, ecosystem, scale = 'fixed', fit.lm = TRUE)  
```

<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr class="header">
<th align="center">Argument</th>
<th align="left">Description</th>
</tr>
</thead>
<tbody>
<tr class="odd">
<td align="center"><code>output</code></td>
<td align="left">the output of <code>mf</code>.</td>

</tr>
<tr class="even">
<td align="center"><code>ecosystem</code></td>
<td align="left"><code>'single'</code> or <code>'multiple'</code>. MF measures in a single ecosystem or multiple ecosystems.</td>

</tr>
<tr class="odd">
<td align="center"><code>scale</code></td>
<td align="left">Are scales shared across all facets (the default, <code>"fixed"</code>), or do they vary across rows (<code>"free_x"</code>), columns (<code>"free_y"</code>), or both rows and columns (<code>"free"</code>)?</td>

</tr>
<tr class="even">
<td align="center"><code>fit.lm</code></td>
<td align="left">If <code>TRUE</code>, the default, fitted line of linear model and estimated of slope are displayed.</td>


</tbody>
</table>


The `ggmf()` function is a wrapper around the `ggplot2` package to display the relationship between species richness and multi-functionality when <code>ecosystem = 'single'</code>, or to display the relationship between the decomposition of species richness and multi-functionality (MF alpha vs. species alpha, MF beta vs. species beta, and MF gamma vs. species gamma) when <code>ecosystem = 'multiple'</code> using a single line of code. The resulting object is of class `"ggplot"` or `"list"` containing two `ggplot` objects, so it can be manipulated using the `ggplot2` tools. Users can visualize the output with or without the fitted line of linear model by setting the parameter <code>**fit.lm**</code>:


We first display multi-functionality in single ecosystem with and without the fitted line.
```{r, fig.align='center', fig.height=8, fig.width=6}
ggmf(output1, ecosystem = 'single', scale = 'fixed', fit.lm = TRUE)
```


```{r, fig.align='center', fig.height=8, fig.width=6}
ggmf(output1, ecosystem = 'single', scale = 'fixed', fit.lm = FALSE)  
```



And then, display decomposition of multi-functionality in multiple ecosystems with and without the fitted line.
```{r, fig.align='center', fig.height=10, fig.width=6}
ggmf(output2, ecosystem = 'multiple', scale = 'fixed', fit.lm = TRUE)$uncorrelated
ggmf(output2, ecosystem = 'multiple', scale = 'fixed', fit.lm = TRUE)$correlated
```

```{r, fig.align='center', fig.height=10, fig.width=6}
ggmf(output2, ecosystem = 'multiple', scale = 'fixed', fit.lm = FALSE)$uncorrelated
ggmf(output2, ecosystem = 'multiple', scale = 'fixed', fit.lm = FALSE)$correlated
```

